Design and Implementation of Probabilistic Programming Language Anglican

### 2017-02-22 paper/paper.tex
minor
update

\subsection*{Contributions} This paper brings the following
major contributions:
\begin{itemize}
    \item Design and implementation of a probabilistic programming language
        Anglican involving tight bilateral integration with a general-purpose
        programming language.
    \item Techniques for efficient and compact implementation of inference
        algorithms, such as representation of inference results as a lazy
        sequence of samples and a novel scheme for addressing of checkpoints.
    \item Anglican's novel representation of random processes 
            from statistics and machine learning, such as Beta-Binomial 
            process. The representation is
            \emph{stateless} and seamlessly integrates into pure functional 
            computation. This contrasts with typical \emph{stateful}
            implementations of these processes in other probabilistic 
            programming languages.
%    \item Specification of probabilistic forms in Anglican which facilitates 
%        separation of probabilistic programs from inference algorithms.
\end{itemize}

@@

The purpose of probabilistic forms \texttt{sample} and
\texttt{observe} is to interrupt deterministic computation and
transfer control to the inference algorithm.  Practically, this
is achieved by  returning a \textbf{checkpoint} instead of
invoking a continuation in a CPS-transformed program.  
Checkpoints are values of the
corresponding record types
(\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}sample} or
\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}observe})
in Clojure. The types contain fields specific to each form, as well as the
continuation; calling the continuation resumes the computation.
\begin{lstlisting}[style=default]
=> (cps-of-expression '(sample dist) 'cont)
(->sample dist cont $state)
=> (cps-of-expression '(observe dist v) 'cont)
(->observe dist v cont $state)
\end{lstlisting}
Here \texttt{->sample} and \texttt{->observe} in the
CPS-transformed expressions are constructors for Clojure
records, and they take values of their fields as arguments.

@@

\texttt{anglican.runtime} is the Anglican runtime library. For
convenience, it exposes common mathematical functions
(\texttt{abs}, \texttt{floor}, \texttt{sin}, \texttt{log},
\texttt{exp}, etc.), but most importantly, it provides
definitions of common distributions. Each distribution object implements a
distribution interface with the \texttt{sample*} and \texttt{observe*} methods;
this interface is defined using Clojure's protocol mechanism
(\texttt{anglican.{\linebreak[0]}runtime/{\linebreak[0]}distribution}).
The \texttt{sample*} \textit{method} returns a random sample and
roughly corresponds to the \texttt{sample} checkpoint,
the \texttt{observe*} \textit{method} returns the log probability
of the value and roughly corresponds to the \texttt{observe}
checkpoint. The methods can be, and sometimes are called from
handlers of the corresponding checkpoints, but do not have to
be. For example, in LMH either the \texttt{sample*} or
the \texttt{observe*} \textit{method} is called for a
\texttt{sample} checkpoint, depending on whether the value is
drawn or reused.

@@

In addition to distributions, Anglican provides \textit{random processes},
which define sequences of random variables that are not independent and
identically distributed. Random processes are similar to the so called
`exchangeable random primitives' in Church \cite{GMR+08} and
Venture \cite{MSP14}. However, random sequences generated by Anglican random 
processes are not required to be exchangeable. Random processes are
defined using the \texttt{defproc} macro and implement the
\texttt{anglican.runtime/{\linebreak[0]}random-process} protocol. This
protocol has two methods
\begin{itemize}
\item \texttt{produce}, which returns the distribution on the next
random variable in the sequence, and
\item \texttt{absorb}, which incorporates the value for the next random
variable and returns an updated random process.
\end{itemize}
 As an example, here is a definition of a beta-Bernoulli process, in which
each random variable is distributed according to a Bernoulli distribution with
an unknown parameter that is drawn from a beta distribution:
\begin{lstlisting}[style=default]
(defproc beta-bernoulli
  "beta-Bernoulli process"
  [a b]
  (produce [this] (bernoulli (/ a (+ a b))))
  (absorb [this x]
    (case x
      0 (beta-bernoulli a (inc b))
      1 (beta-bernoulli (inc a) b))))
\end{lstlisting}
Unlike typical implementations of exchangeable random processes,
Anglican's random processes do not have mutable state. The
\texttt{produce} and \texttt{absorb} methods are deterministic
and functional, and therefore do not have corresponding special
forms in Anglican. A sequence of random values can be generated
using a recursive loop in which \texttt{absorb} returns the
updated process for the next iteration. For example:
\begin{lstlisting}[style=default]
(defm sample-beta-binomial [n a b]
  (loop [process (beta-bernoulli a b)
         values []]
    (if (= (count values) n)
      values
      (let [dist (produce process)
            value (sample dist)]
      (recur (absorb process value)
             (conj values value))))))
\end{lstlisting}
Similarly, random processes can also be used to recursively
observe a sequence of values:
\begin{lstlisting}[style=default]
(loop [process (beta-bernoulli 1.0 1.0)
       [value0 & values0] values] 
  (let [dist (produce process)]
    (observe dist value0)
    (recur (absorb process value0) values0)))
\end{lstlisting}

@@

\begin{lstlisting}[style=default]
(defn towers-of-hanoi [n from to via]
  (when (not= n 1)
      (towers-of-hanoi (dec n) from via to)
      (towers-of-hanoi (dec n) via to from)))
\end{lstlisting}

@@

The other comparison must be made between Anglican and another
probabilistic programming environment. WebPPL~\cite{GS15} is
similar to Anglican in expressivity.  WebPPL runs on top of
\href{http://v8project.blogspot.com/}{V8}, a high performance
JavaScript engine featuring a native compiler.
Table~\ref{tbl:webppl} shows results for a few models from
\href{https://github.com/probmods/webppl}{WebPPL distribution},
as well as from
\href{https://bitbucket.org/probprog/anglican}{Anglican test
suite}. For the comparison, we re-implemented WebPPL models in
Anglican, and Anglican models in WebPPL, following as close as
possible the original implementations, and run the same
inference algorithm (called MCMC in WebPPL, LMH in Anglican)
with the same number of iterations. Anglican outperforms WebPPL
on all models. This can be attributed both to a better
transformed representation of probabilistic programs and to more
efficient architecture of the inference engine.
\begin{table}
    \centering
    \begin{tabular}{r|S[table-format=2.1]|S[table-format=2.1]}
        {\bf model}  & {\bf Anglican} & {\bf WebPPL} \\ \hline
        Latent Dirichlet allocation & 22.2 & 31.5 \\ 
        Linear regression  & 6.1 & 8.5 \\
        Logistic regression & 8.4 & 10.4 \\
        Multivariate regression & 6.9 & 14.2 \\
        Simple branching & 2.4 & 5.7 \\
        Hidden Markov model & 8.7 & 10.2
    \end{tabular}

>>>>

\subsection*{Contributions} This paper brings the following
major contributions:
\begin{itemize}
    \item Design and implementation of a probabilistic programming language
        Anglican involving tight bilateral integration with a general-purpose
        programming language.
    \item Techniques for efficient and compact implementation of inference
        algorithms, such as representation of inference results as a lazy
        sequence of samples and a novel scheme for addressing of checkpoints.
    \item Anglican's novel representation of random processes 
            from statistics and machine learning, such as beta-Bernoulli
            process. The representation is
            \emph{stateless} and seamlessly integrates into pure functional 
            computation. This contrasts with typical \emph{stateful}
            implementations of these processes in other probabilistic 
            programming languages.
%    \item Specification of probabilistic forms in Anglican which facilitates 
%        separation of probabilistic programs from inference algorithms.
\end{itemize}

@@

The purpose of probabilistic forms \texttt{sample} and
\texttt{observe} is to interrupt deterministic computation and
transfer control to the inference algorithm.  Practically, this
is achieved by  returning a \textbf{checkpoint} instead of
invoking a continuation in a CPS-transformed program.  
Checkpoints are represented by values of the
corresponding record types
(\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}sample} or
\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}observe})
in Clojure. The types contain fields specific to each form, as well as the
continuation; calling the continuation resumes the computation.
\begin{lstlisting}[style=default]
=> (cps-of-expression '(sample dist) 'cont)
(->sample dist cont $state)
=> (cps-of-expression '(observe dist v) 'cont)
(->observe dist v cont $state)
\end{lstlisting}
Here \texttt{->sample} and \texttt{->observe} in the
CPS-transformed expressions are constructors for Clojure
records, and they take values of their fields as arguments.

@@

\texttt{anglican.runtime} is the Anglican runtime library. For
convenience, it exposes common mathematical functions
(\texttt{abs}, \texttt{floor}, \texttt{sin}, \texttt{log},
\texttt{exp}, etc.), but most importantly, it provides
definitions of common distributions. Each distribution object implements a
distribution interface with the \texttt{sample*} and \texttt{observe*} methods;
this interface is defined using Clojure's protocol mechanism
(\texttt{anglican.{\linebreak[0]}runtime/{\linebreak[0]}distribution}).
The \texttt{sample*} \textit{method} returns a random sample and
roughly corresponds to the \texttt{sample} checkpoint.
The \texttt{observe*} \textit{method} returns the log probability
of the value and roughly corresponds to the \texttt{observe}
checkpoint. The methods can be, and sometimes are called from
handlers of the corresponding checkpoints, but do not have to
be. For example, in LMH either the \texttt{sample*} or
the \texttt{observe*} \textit{method} is called for a
\texttt{sample} checkpoint, depending on whether the value is
drawn or reused.

@@

In addition to distributions, Anglican provides \textit{random processes},
which define sequences of random variables that are not independent and
identically distributed. Random processes are similar to the so called
`exchangeable random primitives' in Church \cite{GMR+08} and
Venture \cite{MSP14}. However, random sequences generated by Anglican random 
processes are not required to be exchangeable. Random processes are
defined using the \texttt{defproc} macro and implement the
\texttt{anglican.runtime/{\linebreak[0]}random-process} protocol. This
protocol has two methods
\begin{itemize}
\item \texttt{produce}, which returns the distribution on the next
random variable in the sequence, and
\item \texttt{absorb}, which incorporates the value for the next random
variable and returns an updated random process.
\end{itemize}
 As an example, here is a definition of a beta-Bernoulli process, in which
each random variable is distributed according to a Bernoulli distribution with
an unknown parameter that is drawn from a beta distribution:
\begin{lstlisting}[style=default]
(defproc beta-bernoulli
  "beta-Bernoulli process"
  [a b]
  (produce [this] (bernoulli (/ a (+ a b))))
  (absorb [this x]
    (case x
      0 (beta-bernoulli a (inc b))
      1 (beta-bernoulli (inc a) b))))
\end{lstlisting}
Unlike typical implementations of exchangeable random processes,
Anglican's random processes do not have mutable state. The
\texttt{produce} and \texttt{absorb} methods are deterministic
and functional, and therefore do not have corresponding special
forms in Anglican. A sequence of random values can be generated
using a recursive loop in which \texttt{absorb} returns the
updated process for the next iteration. For example:
\begin{lstlisting}[style=default]
(defm sample-beta-binomial [n a b]
  (loop [process (beta-bernoulli a b)
         values []]
    (if (= (count values) n)
      values
      (let [dist (produce process)
            value (sample dist)]
        (recur (absorb process value)
               (conj values value))))))
\end{lstlisting}
Similarly, random processes can also be used to recursively
observe a sequence of values:
\begin{lstlisting}[style=default]
(loop [process (beta-bernoulli 1.0 1.0)
       [value0 & values0] values] 
  (if (= (count values0) 0)
    nil
    (let [dist (produce process)]
      (observe dist value0)
      (recur (absorb process value0) values0))))
\end{lstlisting}

@@

\begin{lstlisting}[style=default]
(defn towers-of-hanoi [n from to via]
  (when (not= n 1) ;; return nil if n = 1.
      (towers-of-hanoi (dec n) from via to)
      (towers-of-hanoi (dec n) via to from)))
\end{lstlisting}

@@

The other comparison must be made between Anglican and another
probabilistic programming environment. WebPPL~\cite{GS15} is
similar to Anglican in expressiveness.  WebPPL runs on top of
\href{http://v8project.blogspot.com/}{V8}, a high performance
JavaScript engine featuring a native compiler.
Table~\ref{tbl:webppl} shows results for a few models from
\href{https://github.com/probmods/webppl}{WebPPL distribution},
as well as from
\href{https://bitbucket.org/probprog/anglican}{Anglican test
suite}. For the comparison, we re-implemented WebPPL models in
Anglican, and Anglican models in WebPPL, following as close as
possible the original implementations, and run the same
inference algorithm (called MCMC in WebPPL, LMH in Anglican)
with the same number of iterations. Anglican outperforms WebPPL
on all models. This can be attributed both to a better
transformed representation of probabilistic programs and to more
efficient architecture of the inference engine.
\begin{table}
    \centering
    \begin{tabular}{r|S[table-format=2.1]|S[table-format=2.1]}
        {\bf model}  & {\bf Anglican} & {\bf WebPPL} \\ \hline
        Latent Dirichlet allocation & 22.2 & 31.5 \\ 
        Linear regression  & 6.1 & 8.5 \\
        Logistic regression & 8.4 & 10.4 \\
        Multivariate regression & 6.9 & 14.2 \\
        Simple branching & 2.4 & 5.7 \\
        Hidden Markov model & 8.7 & 10.2
    \end{tabular}
